---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(scales)
```

## Loading and preprocessing the data
```{r read_data}
unzip("activity.zip")
active_data <- read.csv("activity.csv", header = TRUE, stringsAsFactors=FALSE)
active_data$date <-  as.POSIXct(active_data$date, format="%Y-%m-%d")
dayofweek <- weekdays(active_data$date)
active_data <- cbind(active_data, dayofweek)
summary(active_data)
head(active_data)
```


## What is mean total number of steps taken per day?

### Calculate the number of steps per day
```{r daily_step}
all_steps <- active_data %>% group_by(date) %>% summarise (daily_steps = sum(steps,na.rm=TRUE))
head(all_steps)
```

### Plot number of steps per day
```{r plot_steps}
barplot(height = all_steps$daily_steps, names.arg=all_steps$date, cex.names=0.60, las=3, col = "green")
#steps_median <- median(all_steps$daily_steps, na.rm = TRUE)
#steps_mean <- round(mean(all_steps$daily_steps, na.rm = TRUE),digits = 0)
steps_median <- median(all_steps$daily_steps, na.rm = TRUE)
steps_mean <- round(mean(all_steps$daily_steps, na.rm = TRUE),digits = 0)
abline(h=steps_median,lwd=3, col="blue")
abline(h=steps_mean,lwd=3, col="red")
legend("topleft", c("median", "mean"), fill = c("blue", "red"))
#legend("top", c("median", "mean"), c(steps_median, steps_mean), fill = c("blue", "red"))
```

### Histogram of the total number of steps taken each day
#### (Removed all days with no activity)
```{r plot_hist}
all_steps <- filter(all_steps, all_steps$daily_steps > 0)
hist(all_steps$daily_steps, breaks = nrow(all_steps)/3,  main = "Total number of steps taken each day", 
     xlab = "Total steps taken each day", col = "blue")
abline(v=median(all_steps$daily_steps),lty=5, lwd=2, col="black")
legend(legend="median","topleft",lty=5,lwd=2)
```

### Calculate and report the mean and median of the total number of steps taken per day
```{r mean/median}
steps_mean <- as.integer(mean(all_steps$daily_steps, na.rm = TRUE))
steps_median <- as.integer(median(all_steps$daily_steps, na.rm = TRUE))
```
The average (mean) number of steps taken per day is `r steps_mean` and the median number of steps per day is `r steps_median`.  


## What is the average daily activity pattern?
### Time Series plot showing daily activity pattern
```{r time_plot}
daily_activity <- active_data %>% group_by(interval) %>% summarise(daily_average=mean(steps, na.rm = TRUE))
plot(daily_activity$interval, daily_activity$daily_average, type = "l", col="purple", lwd = 2, xlab="Interval",
     ylab="Average number of steps", main="Average Daily Activiy Pattern")
```

### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r max_interval}
max_interval <- filter(daily_activity,daily_average==max(daily_average))
```

The interval containing the max value is "`r max_interval[1,1]`".  The value for that interval is "`r round(max_interval[1,2],2)`" steps. 



## Imputing missing values
### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r missing_values}
missingvalues <- sum(is.na(active_data$steps))
```
The total number of missing values in the dataset is `r missingvalues`.

### Devise a strategy for filling in all of the missing values in the dataset.
Create average template for all interval of the day.  Use the template to fill in any blanks for missing values.
```{r create_avg_template}
avg_temp <-  daily_activity$daily_average[match(active_data$interval, daily_activity$interval)]
active_data_imputed <- transform(active_data, steps = ifelse(is.na(active_data$steps), 
                                                             yes = avg_temp, no = active_data$steps))
imputed_total_steps <- active_data_imputed %>% group_by(date) %>% summarise (daily_steps = sum(steps))
```

### Make a histogram of the total number of steps taken each day
```{r plot_impute}
hist(imputed_total_steps$daily_steps, col = "gray", xlab = "Total steps per day", 
     main = "Total number of steps taken each day", breaks = nrow(imputed_total_steps)/3)
abline(v=median(imputed_total_steps$daily_steps),lty=5, lwd=2, col="red")
legend(legend="median","topleft",lty=5,lwd=2, col = "red")
```

### Calculate the mean and median total number of steps taken per day.
```{r impute_mean_median}
mean(imputed_total_steps$daily_steps)
median(imputed_total_steps$daily_steps)
```


## Are there differences in activity patterns between weekdays and weekends?
